---
title: "AN588_Boots_imcarlyy"
author: "Carla Rojas"
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    toc: true
    toc_depth: 2
date: "2025-04-14"
---
```{r}
install.packages("ggplot2", repos = "https://cloud.r-project.org/")
```

# Bootstrapping Standard Errors and CIs for Linear Models 

When we initially discussed the central limit theorem and confidence intervals, we showed how we could use bootstrapping to estimate standard errors and confidence intervals around certain parameter values, like the mean. Using bootstrapping, we could also do the same for estimating standard errors and CIs around regression parameters, such as β  coefficients.

## [1] Using the “KamilarAndCooperData.csv” dataset, run a linear regression looking at log(HomeRange_km2) in relation to log(Body_mass_female_mean) and report your β coeffiecients (slope and intercept).

```{r}
library(curl)
f <- curl("https://raw.githubusercontent.com/fuzzyatelin/fuzzyatelin.github.io/refs/heads/master/AN588_Spring25/KamilarAndCooperData.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = TRUE)
head(d)
```

```{r}
m1 <- lm(data=d, formula = log(HomeRange_km2) ~ log(Body_mass_female_mean))
summary(m1)
```

Our coefficient report: 
Slope β coeffiecients: 1.03643
Intercept β coeffiecients: -9.44123 

## [2] Then, use bootstrapping to sample from your data 1000 times with replacement, each time fitting the same model and calculating the same coefficients. This generates a sampling distribution for each β coefficient.

```{r}
set.seed(123)  # For reproducibility (this is optional guys but I found it to be helpful)

boot_results <- data.frame(intercept = numeric(0), slope = numeric(0))

for (i in 1:1000) {
  bootsample <- d[sample(nrow(d), replace = TRUE), ]
  model <- lm(log(HomeRange_km2) ~ log(Body_mass_female_mean), data = bootsample)
  coefs <- coef(model)
  boot_results[i, ] <- coefs  # cleaner indexing instead of rbind (faster too)
}

```

### Estimate the standard error for each of your β coefficients as the standard deviation of the sampling distribution from your bootstrap and determine the 95% CI for each of your β coefficients based on the appropriate quantiles from your sampling distribution.

Estimation of SE: 
```{r}
boot_se <- apply(boot_results, 2, sd)
boot_se
```
Determining the 95% CIs: 
```{r}
boot_ci <- apply(boot_results, 2, quantile, probs = c(0.025, 0.975))
boot_ci
```

```{r}
# Standard errors from lm()
lm_se <- summary(m1)$coefficients[, "Std. Error"]

# 95% Confidence intervals from lm()
lm_ci <- confint(m1)

# Combine into a comparison table
comparison <- data.frame(
  Coef = c("Intercept", "Slope"),
  SE_lm = round(lm_se, 4),
  SE_boot = round(boot_se, 4),
  CI_lm_lower = round(lm_ci[, 1], 4),
  CI_lm_upper = round(lm_ci[, 2], 4),
  CI_boot_lower = round(boot_ci[1, ], 4),
  CI_boot_upper = round(boot_ci[2, ], 4)
)
comparison
```
### How does the former compare to the SE estimated from your entire dataset using the formula for standard error implemented in lm()?

The bootstrapped SE's for both the intercept and slope are very similar to those estimated using the standard lm() approach. However, the bootstrapped SE seems to be slighty lower, showing a little bit more of variablitity than what is assumed under the standard approach.

### How does the latter compare to the 95% CI estimated from your entire dataset?

```{r}
# Finding the 95% CIs estimated from the original model:
og_CIs <- confint(m1)
og_CIs
```

The 95% CIs from `lm()` are a bit more narrow than the ones from bootstrapping. That makes sense because `lm()` uses math formulas based on assumptions, while bootstrapping just watches what happens when we resample the real data. The bootstrap CIs give us a slightly more cautious (but possibly more realistic) sense of uncertainty.

## EXTRA CREDIT

Write a FUNCTION that takes as its arguments a dataframe, “d”, a linear model, “m” (as a character string, e.g., “logHR~logBM”), a user-defined confidence interval level, “conf.level” (with default = 0.95), and a number of bootstrap replicates, “n” (with default = 1000). Your function should return a dataframe that includes: beta coefficient names; beta coefficients, standard errors, and upper and lower CI limits for the linear model based on your entire dataset; and mean beta coefficient estimates, SEs, and CI limits for those coefficients based on your bootstrap.

```{r}
bootstrap_lm_summary <- function(d, m, conf.level = 0.95, n = 1000) {
  # Convert string to formula
  formula <- as.formula(m)
  
  # Fit the original model
  original_model <- lm(formula, data = d)
  original_summary <- summary(original_model)
  original_coef <- coef(original_model)
  original_se <- original_summary$coefficients[, "Std. Error"]
  original_ci <- confint(original_model, level = conf.level)
  
  # Set up bootstrap
  boot_results <- matrix(NA, nrow = n, ncol = length(original_coef))
  colnames(boot_results) <- names(original_coef)
  
  for (i in 1:n) {
    bootsample <- d[sample(nrow(d), replace = TRUE), ]
    model <- lm(formula, data = bootsample)
    boot_results[i, ] <- coef(model)
  }
  
  # Bootstrap summaries
  boot_means <- apply(boot_results, 2, mean)
  boot_ses <- apply(boot_results, 2, sd)
  alpha <- (1 - conf.level) / 2
  boot_cis <- apply(boot_results, 2, quantile, probs = c(alpha, 1 - alpha))
  
  # Create comparison dataframe
  result <- data.frame(
    Coefficient = names(original_coef),
    Beta_lm = round(original_coef, 4),
    SE_lm = round(original_se, 4),
    CI_lm_lower = round(original_ci[, 1], 4),
    CI_lm_upper = round(original_ci[, 2], 4),
    Beta_boot = round(boot_means, 4),
    SE_boot = round(boot_ses, 4),
    CI_boot_lower = round(boot_cis[1, ], 4),
    CI_boot_upper = round(boot_cis[2, ], 4)
  )
  
  return(result)
}
```

This is how I would call this function on my dataset: 
```{r}
bootstrap_lm_summary(
  d = d,
  m = "log(HomeRange_km2) ~ log(Body_mass_female_mean)",
  conf.level = 0.95,
  n = 1000
)

```

## EXTRA EXTRA CREDIT

Graph each beta value from the linear model and its corresponding mean value, lower CI and upper CI from a bootstrap as a function of number of bootstraps from 10 to 200 by 10s. HINT: the beta value from the linear model will be the same for all bootstraps and the mean beta value may not differ that much!
library(ggplot2)

```{r}
library(ggplot2)
bootstrap_trace <- function(d, m, conf.level = 0.95) {
  formula <- as.formula(m)
  original_model <- lm(formula, data = d)
  original_coefs <- coef(original_model)
  
  n_vals <- seq(10, 200, by = 10)
  results <- data.frame()

  for (n in n_vals) {
    boot_mat <- replicate(n, {
      bootsample <- d[sample(nrow(d), replace = TRUE), ]
      coef(lm(formula, data = bootsample))
    })
    boot_mat <- t(boot_mat)
    boot_means <- apply(boot_mat, 2, mean)
    boot_cis <- apply(boot_mat, 2, quantile, probs = c((1 - conf.level)/2, 1 - (1 - conf.level)/2))
    
    for (name in names(original_coefs)) {
      results <- rbind(results, data.frame(
        n = n,
        Coefficient = name,
        Original = original_coefs[name],
        Mean = boot_means[name],
        CI_lower = boot_cis[1, name],
        CI_upper = boot_cis[2, name]
      ))
    }
  }

  ggplot(results, aes(x = n, y = Mean, color = Coefficient)) +
    geom_line() +
    geom_ribbon(aes(ymin = CI_lower, ymax = CI_upper, fill = Coefficient), alpha = 0.2, color = NA) +
    geom_hline(aes(yintercept = Original, linetype = Coefficient), data = unique(results[, c("Coefficient", "Original")]), color = "black", linetype = "dashed") +
    labs(title = "Stability of Bootstrap Estimates by Sample Size",
         x = "Number of Bootstraps",
         y = "Beta Coefficient Estimate") +
    theme_minimal()
}
```

This is how I would call it: 
```{r}
bootstrap_trace(d, "log(HomeRange_km2) ~ log(Body_mass_female_mean)")
```
